Lawvere-Tierney Sheafification\\ in Homotopy Type Theory

### 2015-01-19 paper/sheaf_lics.tex
minor changes

Moreover the following diagram commute
$$\xymatrix{
  \sum_{x:X} (f x = y) \ar[r]^\eta \ar[d]_\gamma & \modal \left(\sum_{x:X}  (f x = y)\right) \ar@{=}[dl]\\
  \sum_{x:\modal X} (\modal f x = \eta_Y y) &
}$$
where $\pi_1 \circ \gamma = \eta_X \circ \pi_1$, and $\pi_2 \circ
\gamma$ is given by the modalisation of paths.
\end{prop}
\begin{proof}
% \nt{ça a du sens de mettre une preuve, non ? c'est non trivial et
%   nouveau par rapport aux travaux sur les modalités. Ou alors c'est
%   la preuve de Shulman qui tu n'as pas faite. En tout cas, il faut
%   clarifier.}

@@

Another example of left exact modality for every level $p$ is the
double negation modality,
$\modal_{\lnot\lnot} T = (T \to \zero) \to \zero$.
%
This modality enables to use
classical reasonings in the corresponding reflective subuniverse, but
every type in the new universe is collapsed to an $\HProp$, ending up
with propositional classical logic.

@@

A fundamental property in the construction of the sheafification is
the fact that the type of modal types at level $n$ defines a sheaf. 
%
This property initially holds in the topos theoretic construction, but
it requires much work here, as we have to deal with equalities. In
particular, this property uses in particular the specific definition
of separated types and sheaves, which shows that we had actually no
choice in their statement.

@@

We begin by proving that $\modal_{n+1}$ is compatible with $\modal_n$
on lower homotopy types.
\begin{prop}
  If $T:\Type_n$, then $\modal_{n+1}T = \modal_n T$.
\end{prop}
\begin{proof}
  We prove it by induction on $n$~:
  \begin{itemize}
  \item For $n=-1$: Let $T:\HProp$. Then
    \begin{align*}
      \modal_{0} T &= \sum_{u:T \to \Type_n^\modal} \modal_{-1}\left\| \sum_{a:T} 
      (\lambda t,~\modal_{-1} (a=t)) = u\right\| \\
      &= \sum_{u:T \to \Type_n^\modal} \modal_{-1}\left( \sum_{a:T} 
      (\lambda t,~\modal_{-1} (a=t)) = u\right)
    \end{align*}
    because the type inside the truncation is already in $\HProp$.
    Now, let define $\phi : \modal_{-1} T \to \modal_0T$ by
    $$\phi t = (\lambda t',\, \modal_{-1}(t=t') ; \eta
    (t;\mathrm{idpath}))$$
    and $\psi : \modal_0T\to \modal_{-1} T$ by obtaining the
    witness $a:T$, and letting
    $$\psi (u;x) = \eta_T a.$$
    These two maps form an equivalence.
  \item Suppose now that $\modal_{n+1}$ is compatible with all $\modal_k$ on
    lower homotopy types. Let $\modal_{n+2}$ be as above, and let
    $T:\Type_{n+1}$. Then
    \begin{align*}
      \modal_{n+2} T &= \sum_{u:T \to \Type_{n+1}^\modal} \modal_{-1}\left\| \sum_{a:T} 
            (\lambda t,~\modal_{n+1} (a=t)) = u\right\| \\
                     &= \sum_{u:T \to \Type_{n+1}^\modal} \modal_{-1}\left\| \sum_{a:T} 
            (\lambda t,~\modal_{n} (a=t)) = u\right\|
    \end{align*}
    because $\modal_{n+1}$ is compatible with $\modal_{n}$, and
    $(a=t):\Type_n$.

    It lefts to prove that $u:T\to\Type_n^\modal$, \ie that for all
    $t:T$, $\IsType n (u\, t)$.
    But for any truncation index $p$, the predicate $\IsType p:\HProp$
    is a a sheaf, so we can get rid of $\modal_{-1}$, and pick up what
    is inside the truncation. Thus, we know that for all $t:T$, $u\, t
    = \modal_n(a=t) : \Type_n$.
  \end{itemize}
\end{proof}

@@

The last step is the left-exactness of $\modal_{n+1}$. Let
$T:\Type_{n+1}$, and $x,y:T$. Suppose that $\Contr(\modal_{n+1} T)$.
Thanks to compatibility between $\modal_{n+1}$ and $\modal_n$ for
$n-\Type$, it suffices to show that $\modal_n(x=y)$ is contractible.

@@

Overall, with other files containing technical lemmas, the project
contains 6322 lines, and it could be reduced a bit by improving the
way Coq tries to rewrite and apply lemmas automatically. 
%
This constitute a significant amount of work but the part dedicated to
sheaves and sheafification is only 4000, which seems quite reasonable
and is encouraging, because it suggests that homotopy type theory
provides a convenient tool to formalize some part of the theory of
higher topoi. As a comparison, this \TeX\xspace file contains
approximatively 1800 lines.

@@

In our formalization of sheafification defined
Section~\ref{sec:sheafification} (done in 
\texttt{sheaf\_induction.v}), two very technical facts required to
show that $\square$ is a modality have been
admitted. Nevertheless, we are confident that they can be proven and
have been admitted only by lack of time.
In the proof that diagrams are equal in lemma~\ref{lem:sepiscolim},
the equality of projections is admitted, because if this proof is
obvious on paper, it becomes really technical when formalized.
As a consequence, the fact that $P$ is a colimit of the diagonal
closed diagram is admitted too, but again, this proof should be really
short as it is on paper.

@@

We have chosen a more simple but less satisfactory solution. 
%
When the absence of cumulativity rises an universe issue for a lemma,
we have proven the lemma and then admit the exact same lemma, just to
let Coq infer less constrained universe levels.

An other issue with universe polymorphism lies in their management for
recursive definitions. Indeed, the definition of successive sheafification types is
not allowed:
%
\[ \left\{
    \begin{array}{l}
      \modal_{-1} = \lambda T:\HProp,~ \lnot\lnot T \\
      \displaystyle{\modal_{n+1} = \lambda T:\Type_{n+1}},\\
      \quad \displaystyle{\sum_{u:T \to \Type_n^\modal} \modal_{-1} 
      \left\|
      \sum_{a:T} u = (\lambda t,~\modal_n (a=t))
      \right\|}
    \end{array}
  \right.\]
%
because Coq asks for every $\modal_n$ to have the same universes
levels in the recursive definition. 
%
Thus, the induction step presented in this paper has been formalized,
but the complete induction can not be formalized at the moment. 
%
This restriction in our formalization can not be solved without
generalizing universe polymorphism for recursive definition.
%
% In particular, compoatcumulativity is an axiom, because of
% universes issues.

>>>>

Moreover the following diagram commute
$$\xymatrix{
  \sum_{x:X} (f x = y) \ar[r]^\eta \ar[d]_\gamma & \modal \left(\sum_{x:X}  (f x = y)\right) \ar@{=}[dl]\\
  \sum_{x:\modal X} (\modal f x = \eta_Y y) & }$$
where $\pi_1 \circ \gamma = \eta_X \circ \pi_1$, and
$\pi_2 \circ \gamma$ is given by the modalisation of
paths.\footnote{This proposition comes from on old version of the
  formalization of left exact modalities in the HoTT library, but is
  not present anymore in the library.}
\end{prop}
\begin{proof}
% \nt{ça a du sens de mettre une preuve, non ? c'est non trivial et
%   nouveau par rapport aux travaux sur les modalités. Ou alors c'est
%   la preuve de Shulman qui tu n'as pas faite. En tout cas, il faut
%   clarifier.}

@@

Another example of left exact modality for every level $p$ is the
double negation modality,
$\modal_{\lnot\lnot} T = (T \to \zero) \to \zero$.
%
This modality is consistent as $\modal_{\lnot\lnot} \zero = \zero$ can
not be inhabited.
%
Using the double negation modality enables to use classical reasonings
in the corresponding reflective subuniverse, but every type in the new
universe is collapsed to an $\HProp$, ending up with propositional
classical logic.

@@

A fundamental property in the construction of the sheafification is
the fact that the type of modal types at level $n$ defines a sheaf. 
%
This property initially holds in the topos theoretic construction, but
it requires much work here, as we have to deal with equalities. In
particular, this property uses the specific definition
of separated types and sheaves, which shows that we had actually no
choice in their statement.

@@

We begin by proving that $\modal_{n+1}$ is compatible with $\modal_n$
on lower homotopy types.
\begin{prop}
  If $T:\Type_n$, then $\modal_{n+1}T = \modal_n T$.
\end{prop}
\begin{proof}
  We prove it by induction on $n$~:
  \begin{itemize}
  \item For $n=-1$: Let $T:\HProp$. Then
    \begin{align*}
      \modal_{0} T &= \sum_{u:T \to \Type_n^\modal} \modal_{-1}\left\| \sum_{a:T} 
      (\lambda t,~\modal_{-1} (a=t)) = u\right\| \\
      &= \sum_{u:T \to \Type_n^\modal} \modal_{-1}\left( \sum_{a:T} 
      (\lambda t,~\modal_{-1} (a=t)) = u\right)
    \end{align*}
    because the type inside the truncation is already in $\HProp$.
    Now, let define $\phi : \modal_{-1} T \to \modal_0T$ by
    $$\phi t = (\lambda t',\, \modal_{-1}(t=t') ; \eta
    (t;\mathrm{idpath}))$$
    and $\psi : \modal_0T\to \modal_{-1} T$ by obtaining the
    witness $a:T$ (which is possible because we are trying to inhabit
    a modal proposition), and letting
    $$\psi (u;x) = \eta_T a.$$
    These two maps form an equivalence (the section and retraction are
    trivial because the equivalence is between mere propositions).
  \item Suppose now that $\modal_{n+1}$ is compatible with all $\modal_k$ on
    lower homotopy types. Let $\modal_{n+2}$ be as above, and let
    $T:\Type_{n+1}$. Then
    \begin{align*}
      \modal_{n+2} T &= \hspace{-1em} \sum_{u:T \to \Type_{n+1}^\modal}
                       \hspace{-1em} \modal_{-1}\left\| \sum_{a:T} 
            (\lambda t,~\modal_{n+1} (a=t)) = u\right\| \\
                     &= \hspace{-1em} \sum_{u:T \to
                       \Type_{n+1}^\modal} 
\hspace{-1em} \modal_{-1}\left\| \sum_{a:T} 
            (\lambda t,~\modal_{n} (a=t)) = u\right\|
    \end{align*}
    because $\modal_{n+1}$ is compatible with $\modal_{n}$, and
    $(a=t):\Type_n$.

    It remains to prove that for every $(u,x)$ inhabited the
    $\Sigma$-type above, $u$ is in $T\to\Type_n^\modal$, \ie that for
    every $t:T$, $\IsType n (u\, t)$.  But for any truncation index $p$,
    the predicate $\IsType p:\HProp$ is a a sheaf, so we can get rid
    of $\modal_{-1}$ and of the truncation, which tells us that for
    every 
    $t:T$, $u\, t = \modal_n(a=t) : \Type_n$.
  \end{itemize}
\end{proof}

@@

The last step is the left-exactness of $\modal_{n+1}$. Let
$T:\Type_{n+1}$, and $x,y:T$. Suppose that $\Contr(\modal_{n+1} T)$.
Thanks to compatibility between $\modal_{n+1}$ and $\modal_n$ for
$\Type_n$, it suffices to show that $\modal_n(x=y)$ is contractible.

@@

Overall, with other files containing technical lemmas, the project
contains 6322 lines, and it could be reduced a bit by improving the
way Coq tries to rewrite and apply lemmas automatically. 
%
This constitute a significant amount of work but the part dedicated to
sheaves and sheafification is only 4000, which seems quite reasonable
and is encouraging, because it suggests that homotopy type theory
provides a convenient tool to formalize some part of the theory of
higher topoi. As a matter of comparison, this \TeX\xspace file contains
approximatively 1800 lines.

@@

In our formalization of sheafification defined
Section~\ref{sec:sheafification} (done in 
\texttt{sheaf\_induction.v}), two very technical facts required to
show that $\square$ is a modality have been
admitted. 
\nt{say that it comes from a lack of lemmas on paths between paths in
  the HoTT library}
%
Nevertheless, we are confident that they can be proven and
have been admitted only by lack of time.
In the proof that diagrams are equal in lemma~\ref{lem:sepiscolim},
the equality of projections is admitted, because if this proof is
obvious on paper, it becomes really technical when formalized.
As a consequence, the fact that $P$ is a colimit of the diagonal
closed diagram is admitted too, but again, this proof should be really
short as it is on paper.

@@

We have chosen a more simple but less satisfactory solution. 
%
When the absence of cumulativity rises a universe issue for a lemma,
we have proven the lemma and then admit the exact same lemma, just to
let Coq infer less constrained universe levels.

An other issue with universe polymorphism lies in their management for
recursive definitions. Indeed, the definition of successive sheafification types is
not allowed:
%
\[ \left\{
    \begin{array}{l}
      \modal_{-1} = \lambda T:\HProp,~ \lnot\lnot T \\
      \displaystyle{\modal_{n+1} = \lambda T:\Type_{n+1}},\\
      \quad \displaystyle{\sum_{u:T \to \Type_n^\modal} \modal_{-1} 
      \left\|
      \sum_{a:T} u = (\lambda t,~\modal_n (a=t))
      \right\|}
    \end{array}
  \right.\]
%
because Coq asks for every $\modal_n$ to have the same universes
levels in the recursive definition. 
%
\nt{explain why it is an issue.}
%
Thus, the induction step presented in this paper has been formalized,
but the complete induction can not be formalized at the moment. 
%
This restriction in our formalization can not be solved without
generalizing universe polymorphism for recursive definition.
%
% In particular, compoatcumulativity is an axiom, because of
% universes issues.
